### 2022-10-03 AISTATS2023/paper.tex
added some content

\aistatstitle{Instructions for Paper Submissions to AISTATS 2023}

@@

\begin{abstract}
  The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
  both left and right-hand margins. Use 10~point type, with a vertical
  spacing of 11~points. The \textbf{Abstract} heading must be centered,
  bold, and in point size 12. Two line spaces precede the
  Abstract. The Abstract must be limited to one paragraph.
\end{abstract}

@@

Recently, ML has flourished in critical domains, such as healthcare
and finance. In these areas, we need a combination of accurate
predictions along with meaningful explanations to support them. For
this reason there is an increased interest in Explainable AI (XAI),
the field that provides interpretations about the behavior of complex
black-box models. XAI literature distinguishes between local and
global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the various local explanations into a single interpretable outcome,
usually a number or a plot. If a user wants to get a rough overview
about which features are significant (feature importance) or whether a
particular feature has a positive or negative effect on the output
(feature effect), they should opt for a global explainability
technique. On the other hand, aggregating the individual explanations
for producing a concise global one is vulnerable to
misinterpretations. Under strong interactions and correlations between
features, the global explanation may obfuscate heterogeneous effects
that exist under the hood~\citep{herbinger2022repid}; a phenomenon
called aggregation bias~\citep{mehrabi2021survey}.

Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category
of global explainability methods. The objective of FE is to the
isolate and visualize the impact of a single feature on the
output.~\footnote{FE methods also isolate the effect of a pair of
  features to the output. Combinations of more than two features are
  not usual, because they encounter, among others, visualization
  difficulties.} FE methods suffer from aggregation bias because,
often, the rationale behind the average effect might be unclear. For
example, a feature with zero average effect may indicate that the
feature has no effect on the output or, contrarily, it has a highly
positive effect in some cases and a highly negative in others. There
are three widely-used FE methods; Partial Dependence Plots
(PDP)\citep{friedman2001greedy}, Marignal Plots
(MP)\citep{apley2020visualizing} and Aggregated Local Effects
(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
computing erroneous effects when the input features are (highly)
correlated, which is a frequent scenario in many ML
problems. Therefore, ALE has been established as the state-of-the-art
FE method.

However, ALE faces two crucial limitations, the first concerns ALE
definition and the second one the ALE approximation. Regarding the
definition ALE does not inform the user about the level of
heterogeneous effects hidden behind the average effect (global
explanation). In contrast, in the case of PDP, the heterogeneous
effects can be (partially) spotted by exploring the Individual
Conditional Expectations (ICE)\citep{goldstein2015peeking}. Regarding
the approximation, e.g.~the estimation of ALE from the limited samples
of the training set, ALE requires an important additional step, called
\textit{bin-splitting}. Bin-splitting consists of partitioning the
axis of the feature of interest in a sequence of non-overlapping
intervals and estimating a single effect in each one, from the
population of samples that fall inside. Specifying an approriate
sequence of intervals is of particular importance, since ALE's
interpretation is meaningful only inside each interval
\citep{molnar2022}. However, this crucial step has not raised the
appropriate attention, making the approximation vulnerable to
potential misinterpretations.

In this paper, we extend ALE with a probabilistic component for
measuring the uncertainty of the explanation. The uncertainty of the
global explanation, i.e.~how certain we are that the averaged
explanation is valid if applied to an instance drawn at random,
quantifies the level of heterogeneous effects. Given that ALE's
interpretation (expected effect and uncertainty) is meaningful only
inside each interval, we design a principled framework, where we treat
the bin-splitting step as a data-driven clustering problem, searching
for the optimal splitting given available instances of the training
set. We, also, present a computationally-grounded algorithm for
finding the optimal solution.

\paragraph{Contributions.} The contribution of this paper is NAME, a
feature effect method that:

\begin{itemize}
\item Extends ALE with a probabilistic component to quantify the
  heterogeneous effects behind the global explanation.
\item Automatically extracts regions with similar effects, improving
  the estimation and the interpretability of ALE plots.
\end{itemize}

@@

At Section~\ref{sec:feat-effect-meth}, we present the basic Feature
Effect (FE) methods and we review the ALE
definition\citep{apley2020visualizing}. At
Section~\ref{sec:quant-heter-effects}, we present the techniques for
quantifying the heterogeneous effects and we discuss some of their
limitations. Finally, at Section~\ref{sec:ale-approximation}, we
review the ALE approximation\citep{apley2020visualizing}(add us),
describing why it is vulnerable to a poor approximation due to
inappropriate bin-splitting.

@@

The three well-known feature effect methods are: PDP, MP and ALE. PDP
formulates the FE of the \(s\)-th attribute as an expectation over the
marginal distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
formulates it as an expectation over the conditional
\(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|x_s}[f(x_s,
\mathbf{X}_c)]\). ALE defines the local effect of the \(s\)-th feature
at point \(x_s = z\) as \(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). All the
local explanations at \(z\) are, then, weighted by the conditional
distribution \(p(\xc|x_s=z)\) and are averaged, to produce the
averaged effect \(\mu(z)\). Finally, ALE is the accumulation of the
averaged local effects:

@@

ALE has specific advantages which gain particular value in cases of
correlated input features. In these cases, PDP integrates over
unrealistic instances due to the use of the marginal distribution
\(\mathbf{X}_c \), and MP computes aggregated effects, i.e., imputes
the combined effect of sets of features to a single feature. ALE
manages to resolve both issues, and is therefore the most trustable
method in cases of correlated features.

@@

FE methods pose the question \textit{what is average (global) effect
  on the output, if the value of a specific feature is
  increased/decreased}. After that, it comes naturally to also ask
about \textit{to what extent the individual (local) effects deviate
  from the average}. For this reason, the quantification of the
uncertainty has attracted a lot of interest.

ICE and d-ICE\citep{goldstein2015peeking} provide a set of curves that
are plot on top-of the PDP. Both methods produce one curve for each
instance of the training set;
\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) for ICE and
\(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
\xc^i)\) for d-ICE. The user then visually observes if the (d)-ICE
curves are homogeneous and to what extent they deviate from the
PDP. Some methods try to automate the aforementioned visual
exploration, by grouping (d-)ICE into clusters~\citep{molnar2020model,
  herbinger2022repid, britton2019vine}. Some other approaches, like
H-Statistic\citep{friedman2008predictive}, Greenwell's interaction
index\citep{greenwell2018simple} or SHAP interaction
values\citep{lundberg2018consistent}, quantify the level of
interaction between the input features, with an interaction value. A
strong interaction index is an indicator for the existence of
heterogeneous effects.

The aforementioned approaches are under two limitations; They either
do not quantify the uncertainty of the FE directly or they are based
on PDPs, and, therefore, they are subject to the failure modes of PDPs
in cases of correlated features\citep{baniecki2021fooling}. To the
best of our knowledge, no work exist so far that quantifies the
heterogeneous effects based on the formulation of ALE.

@@

In real ML scenarios, the FE is estimated from the limited instances
of the training set. Therefore, \citep{apley2020visualizing} proposed
dividing the \(s\)-th axis in \(K\) bins and estimating the local
effects in each bin by evaluating the black box-function at the bin
limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards, (cite) proposed the Differential ALE (DALE)
approximation for computing the local effects on the training
instances using auto-differentiation:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} f^s(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost.

Both approximations ask from the user to blindly decide the number of
bins \(K\) for splitting the axis into equally-sized bins, an approach
with crucial limitations. Firstly, it is vulnerable to poor bin-effect
approximations that lead to a misleading ALE plot. Setting \(K\) to a
small value may hide fine-grain effects due to large bins and setting
\(K\) to a high value leads to estimations of many bin-effects from
very limited samples. In general, the user may face contradictory
explanations for differnt \(K\) without a clue to decide which one to
trust. Secondly, as indicated by~\citep{molnar2022}, the ALE's feature
effect can only be interpreted in local regions. Therefore, for better
interpretation, the extend of these regions should be determined
in a data-driven way.

\section{THE NAME METHOD}
\label{sec:NAME-method}

The NAME method extends ALE for quantifying the level of heterogeneous
effects (uncertainty) and automating the bin-splitting step for robust
estimations. At Section~\ref{sec:NAME-definition}, we define NAME,
presenting the component that extends ALE with uncertainty
quantification. In Section~\ref{sec:interval-based-estimation}, we
provide an alternative definition of NAME based on variable-size bins,
providing important remarks and proofs for connecting it with the
initial definition. In Section~\ref{sec:interval-spliting}, we define
the problem of optimal bin-splitting and we provide an algorithmic
solution to the problem.

\subsection{NAME: ALE With Uncertainty Quantification}
\label{sec:NAME-definition}

ALE does not shed light to potential heterogeneous effects that
contribute to the average explanation. Therefore, we extend ALE
definition of Eq.~(\ref{eq:ALE}) with a component that quantifies the
level of heterogeneous effects. We denote \(\mathcal{H}(z)\) the
uncertainty of the local effects at a specific point \(x_s=z\) and we
quantify it as the standard deviation of the local explanations:
\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(z) := \sigma(z) = \sqrt{\mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] }
\end{equation}
\noindent
The uncertainty emerges from the natural characteristics of the
experiment, i.e.,~the feature correlations existent in the data
generating distribution and the implicit interactions of the black-box
function. We also define the accumulated uncertainty at \(x_s\), as
the accumulation of the standard deviation of the local effects along
the axis:
\begin{equation}
  \label{eq:ALE_acc_unc}
  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
\end{equation}
\noindent
NAME method formulates the effect at a specific point \(x_s\) with a
compact entity that consists of the average effect and the uncertainty
\(\mu(z) \pm \sigma(z)\) and visualizes them as a continuous curve
with a confidence region, i.e.
\(f^{\mathtt{METHOD}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
f^{\mathtt{ALE}}_{\sigma}(x_s)\).

\subsection{Interval-Based Formulation of NAME}
\label{sec:interval-based-estimation}

In real scenarios, the estimations are based on the limited instances
of the training set. Estimating \(\mu(z), \sigma(z)\) at the
granularity of a point is impossible, because the probability of
observing a sample inside the interval \([x_s - h, x_s + h]\) tends to
zero, when \(h \to 0\). We define the mean effect (bin-effect) and the
uncertainty (bin-uncertainty) inside an interval \([z_1, z_2)\) as:

@@

\begin{equation}
  \label{eq:var_bin}
  \sigma(z_1, z_2) = \frac{\int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 - z_1}
\end{equation}

%
We also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points that
partition the \(s\)-th feature axis into a sequence \(K\)
variable-size intervals, i.e. \(\mathcal{Z}_s = \{z_0, z_K\}\), where
\(z_0=x_{s,min}\) and \(z_K=x_{s,max}\). Based on the above, we
redefine NAME using an interval-based formulation
\(\tilde{f}^{\mathtt{METHOD}}(x_s):=
\tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) \pm \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:

@@

\textit{If we define the residual \(\rho(z)\) as the
difference between the expected effect at \(x_s\) and the mean
expected effect at the interval, i.e
\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
\(\sigma^2(z_1, z_2)\) equals to the accumulation of the point-wise
variances plus the squared residuals inside the interval:
\begin{equation}
 \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
\end{equation}}
%

@@

\textit{If we define as \(\rho^2(z_{k-1}, z_{k})\) the accumulated residual of
a bin, i.e.
\(= \rho^2(z_{k-1}, z_{k}) = \int_{z_{k-1}}^{z_k} \rho^2(z)\partial
z\), then the two definitions of NAME are equivalent, i.e.,
\(f^{\mathtt{NAME}} = \tilde{f}^{\mathtt{NAME}}\), if and only if the
sequence of limits \(\mathcal{Z}\) is such that
\(\sum_{k=1}^K \rho^2(z_{k-1}, z_{k}) = 0\).}

@@

Theorem 1 decouples the aggregated uncertainty inside a bin, i.e.,
\(\mathcal{H}_{bin}(z_1, z_2) := \sigma^2(z_1, z_2)\), into two
terms. The first term quantifies the accumulated uncertainty due to
the natural characteristics of the experiment, i.e.,
\(\mathcal{H}(z_1, z_2) = \int_{z_1}^{z_2}\mathcal{H}(z) \partial z\),
and the second term adds extra nuisance uncertainty due to limiting
the resolution
\(\mathcal{H}_n(z_1, z_2) := \int_{z_1}^{z_2}\rho^2(z) \partial z
\). Therefore, it holds that:
\begin{equation}
  \label{eq:uncertainty_bin}
  \mathcal{H}_{bin}(z_1, z_2) = \mathcal{H}(z_1, z_2) + \mathcal{H}_n(z_1, z_2)
\end{equation}
%
In Theorem 2 we prove that if we find a sequence of bins, where each
bin adds no nuisance uncertainty, then we will get exactly the same
NAME explanation as in the original definition.

@@

Both approximations are unbiased only under the assumption that the
points are uniformly distributed inside the intervals. Under this
assumption, and due to the law of large numbers, the approximations
become more robust, as the bins become wider. On the other hand, bins
should group local explanations with similar effects and enlarging a
bin is under the danger of adding nuisance uncertainty. Therefore, we
set-up the following optimization problem:
\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
  & \tau_k = 1 - 0.2 \frac{|S_k|}{N} \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
We search for the sequence of intervals \(z_0, \ldots, z_K\) that
minimizes the sum of the cost of each bin. The cost of each bin is the
mean variance \(\sigma^2_k\) scaled by the bin length \(\Delta z_k\)
and discounted by the term \(\tau_k\). The term \(\tau_K\) acts as a
tie-braker in cases of two candidate bins with equal mean variance,
favoring the selection of the bigger bin. The constraint of at least
\(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
\textit{robust} estimation.

@@

The simulation examples, where the data-generating distribution
\(p(\mathbf{X})\) and the predictive function \(f(\cdot)\) are defined
by us, enable the evaluation of competitive approaches against a solid
ground-truth. We follow this common XAI
practice~\citep{aas2021explaining, herbinger2022repid} for providing
secure empirical evidence about the superiority of NAME method.

>>>>

\aistatstitle{Accumulated Regional Effects: An ALE-based method for regional effects with uncertainty quantification}

@@

\begin{abstract}
  This paper is on Feature Effect (FE), a class of global explainability techniques that create a plot with the average effect of a single feature on the output. The task is particularly challenging when the black-box function contains non-linear interactions between correlated features. In these cases, FE should provide, first, the expected effect and, second, the uncertainty of the effect, i.e.~the level of heterogeneous effects due to implicit feature interactions. Accumulated Local Effects (ALE) is a novel method that models accurately the expected effect in cases of correlated features, overcoming limitations of previous approaches. In this paper, we define Accumulated Regional Effects (ARE) a FE method that extends ALE to quantify the heterogeneous effects. We also provide an alternative definition of ARE based on a sequence of interpretable variable-size regions and we formally prove that under some conditions both definitions are equivalent. Using that, we convert the typical bin-splitting step of ALE to a principled optimization problem and we present an algorithm for finding the optimal sequence of regions. We demonstrate through synthetic and real-datasets, that ARE quantifies correctly the average and heterogeneous effects and approximates them robustly by extracting regions with similar-effect.
\end{abstract}

@@

Recently, Machine Learning (ML) has flourished in critical domains,
such as healthcare and finance. In these areas, we need a combination
of accurate predictions along with meaningful explanations to support
them. For this reason there is an increased interest in Explainable AI
(XAI), the field that provides interpretations about the behavior of
complex black-box models. XAI literature distinguishes between local
and global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the various local explanations into a single interpretable outcome,
usually a number or a plot. If a user wants to get a rough overview
about which features are significant (feature importance) or whether a
particular feature has a positive or negative effect on the output
(feature effect), they should opt for a global explainability
technique. On the other hand, aggregating the individual explanations
for producing a concise global one is vulnerable to
misinterpretations. Under strong interactions and correlations between
features, the global explanation may obfuscate heterogeneous effects
that exist under the hood~\citep{herbinger2022repid}; a phenomenon
called aggregation bias~\citep{mehrabi2021survey}.

Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category
of global explainability methods. The objective of FE is to the
isolate and visualize the impact of a single feature on the
output.~\footnote{Often, FE methods also isolate the combined effect
  of a pair of features to the output. Combinations of more than two
  features are not usual, because they encounter, among others,
  visualization difficulties.} FE methods suffer from aggregation bias
because, often, the rationale behind the average effect is
unclear. For example, a feature with zero average effect may indicate
no effect on the output or, in contrast, highly positive in some cases
and highly negative in others. There are three widely-used FE methods;
Partial Dependence Plots (PDP)\citep{friedman2001greedy}, Marignal
Plots (MP)\citep{apley2020visualizing} and Aggregated Local Effects
(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
computing erroneous effects when the input features are (highly)
correlated, which is a frequent scenario in many ML cases. Therefore,
ALE has been established as the state-of-the-art FE method.

However, ALE faces two crucial limitations, the first concerns ALE
definition and the second one ALE approximation. Regarding the
definition ALE does not inform the user about the level of
uncertainty, i.e., the heterogeneous effects hidden behing the average
effect due to implicit feature interactions. In contrast, in PDPs the
heterogeneous effects can be (partially) spotted by exploring the
Individual Conditional Expectations
(ICE)\citep{goldstein2015peeking}. Regarding the approximation,
i.e.~the estimation usign the (limited) samples of the training set,
ALE requires the additional \textit{bin-splitting} step. Bin-splitting
consists of partitioning the axis of the feature of interest in a
sequence of non-overlapping intervals (regions) and estimating a
constant effect in each one from the population of samples that fall
inside. Specifying an approriate sequence of regions is of particular
importance, since ALE's interpretation is meaningful only inside each
region \citep{molnar2022}. However, this crucial step has not raised
the appropriate attention, making the approximation vulnerable to
potential misinterpretations.

In this paper, we present Accumulated Regional Effects (ARE) an
extension of ALE for measuring the uncertainty of the explanation,
i.e.~how certain we are that the average explanation is valid if
applied to an instance drawn at random. We also provide an alternative
definition of ARE based on a sequence of constant-effect regions and
we define a metric of the dissimilarity of the two definitions. We
formally prove that under some conditions the dissimilarity is zero
and both definitions are equivalent. We, therefore, set up an
optimization problem where we search for the sequence of region that
minimizes the disimilarity, given enough training points iniside each
region for a robust estimation. Finally, we propose a
computationally-grounded algorithm for finding the optimal solution.

\paragraph{Contributions.} The contribution of this paper are:

\begin{itemize}
\item ARE, a FE method that extends ALE for quantifying the
  heterogeneous effects (uncertainty).
\item An interval-based definition of ARE and a formal proof that
  under some conditions it is equivalent to the initial definition.
\item A principled framework for automatically extracting regions with
  similar effects, improving the estimation and the interpretability
  of ALE plots.
\end{itemize}

@@

In this section, we describe the basic methods for FE and for
uncertainty quantification, focusing on ALE definition. We, then,
review the ALE approximation\citep{apley2020visualizing, gkolemis22},
describing some of its vulnerabilities.

@@

The three well-known feature effect methods are: PDP, MP and ALE. PDP
formulates the FE of the \(s\)-th attribute as an expectation over the
marginal distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
formulates it as an expectation over the conditional
\(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|x_s}[f(x_s,
\mathbf{X}_c)]\). ALE defines the local effect of the \(s\)-th feature
at point \(x_s = z\) as
\(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). All the
local explanations at \(z\) are, then, weighted by the conditional
distribution \(p(\xc|x_s=z)\) and are averaged, to produce the
averaged effect \(\mu(z)\). ALE is the accumulation of the averaged
local effects:

@@

Eq.~(\ref{eq:ALE}) has specific advantages which gain particular value in cases of
correlated input features. In these cases, PDP integrates over
unrealistic instances due to the use of the marginal distribution
\(\mathbf{X}_c \), and MP computes aggregated effects, i.e., imputes
the combined effect of sets of features to a single feature. ALE
manages to resolve both issues, and is therefore the most trustable
method in cases of correlated features.

@@

FE methods reply to the question \textit{what is the average (global)
  effect on the output, if the value of a specific feature is
  increased/decreased}. It comes naturally to also ask \textit{to what
  extent the local effects deviate from the global
  explanation}. Quantifying the uncertainty of the global explanation
has attracted a lot of interest. ICE and
d-ICE\citep{goldstein2015peeking} provide a set of curves that are
plot on top-of the PDP. Both methods produce one curve for each
instance of the training set;
\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) for ICE and
\(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
\xc^i)\) for d-ICE. The user then visually observes if the curves are
homogeneous and to what extent they deviate from the PDP. Some methods
try to automate the aforementioned visual exploration, by grouping
(d-)ICE plots into clusters~\citep{molnar2020model,
  herbinger2022repid, britton2019vine}. Some other approaches, like
H-Statistic\citep{friedman2008predictive}, Greenwell's interaction
index\citep{greenwell2018simple} or SHAP interaction
values\citep{lundberg2018consistent}, quantify the level of
interaction between the input features, with an interaction value. A
strong interaction index is an indicator for the existence of
heterogeneous effects.

The aforementioned approaches are under two limitations; They either
do not quantify the uncertainty of the FE directly or they are based
on PDPs, and, therefore, they are subject to the failure modes of PDPs
in cases of correlated features\citep{baniecki2021fooling}, as we will
show in-depth in Section~\ref{sec:simulation-examples-1}. To the best
of our knowledge, no work exist so far that quantifies the
heterogeneous effects based on the formulation of ALE.

@@

In real ML scenarios, the FE is estimated from the training set
examples. Therefore, \citep{apley2020visualizing} proposed dividing
the \(s\)-th axis in \(K\) bins and estimating the local effects in
each bin by evaluating the black box-function at the bin limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards,~\citep{gkolemis22} proposed the Differential ALE
(DALE) that computes the local effects on the training instances using
auto-differentiation:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} f^s(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost.

Both approximations ask from the user to blindly decide the number of
bins, denoted with \(K\), for splitting the axis into equally-sized
bins, an approach with crucial limitations. Setting \(K\) to a small
value may hide fine-grain effects due to large bins and setting \(K\)
to a high value leads to poor bin-effect estimations from limited
samples. In general, the user may face contradictory explanations for
different \(K\) without a clue to decide which one to trust.

\section{Accumulated Regional Effects (ARE)}
\label{sec:NAME-method}

ARE extends ALE for quantifying the level of heterogeneous effects
(uncertainty) and automates the bin-splitting step for robust
estimations. At Section~\ref{sec:NAME-definition}, we define ARE and
at Section~\ref{sec:interval-based-estimation} we provide an
alternative definition of ARE based on variable-size bins, providing
important remarks and proofs for connecting it with the initial
definition. In Section~\ref{sec:interval-spliting}, we define the
problem of optimal bin-splitting and we propose an algorithmic
solution to the problem.

\subsection{ARE: ALE With Uncertainty Quantification}
\label{sec:NAME-definition}

We extend ALE definition of Eq.~(\ref{eq:ALE}) with a component for
quantifying the uncertainty. We denote as \(\mathcal{H}(z)\) the
uncertainty of the local effects at a specific point \(x_s=z\) and we
quantify it as the standard deviation of the local explanations:
\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(z) := \sigma(z) = \sqrt{\mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] }
\end{equation}
\noindent
The uncertainty emerges from the natural characteristics of the
experiment, i.e.,~the feature correlations existent in the data
generating distribution and the implicit interactions of the black-box
function. We also define the accumulated uncertainty at \(x_s\), as
the accumulation of the standard deviation of the local effects along
the axis:
\begin{equation}
  \label{eq:ALE_acc_unc}
  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
\end{equation}
\noindent
ARE method formulates the effect at a specific point \(x_s\) with a
compact tuple that consists of the average effect and the uncertainty
\((\mu(z), \sigma(z))\) and visualizes them as a continuous curve
with a confidence region, i.e.
\(f^{\mathtt{ARE}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
f^{\mathtt{ALE}}_{\sigma}(x_s)\).

\subsection{Interval-Based Formulation of ARE}
\label{sec:interval-based-estimation}

In real scenarios, the estimations are based on the limited instances
of the training set. Estimating \(\mu(z), \sigma(z)\) at the
granularity of a point is impossible, because the probability of
observing a sample inside the region \([x_s - h, x_s + h]\) tends to
zero, when \(h \to 0\). We define the region-effect and the
region-uncertainty inside the region \([z_1, z_2)\) as:

@@

\begin{equation}
  \label{eq:var_bin}
  \sigma(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 - z_1}}
\end{equation}

%
We also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points
that partition the \(s\)-th feature axis into \(K\) variable-size
intervals, i.e. \(\mathcal{Z}_s = \{z_0, z_K\}\), where
\(z_0=x_{s,min}\) and \(z_K=x_{s,max}\). Based on the above, we
redefine ARE using an interval-based formulation
\(\tilde{f}^{\mathtt{ARE}}(x_s):=
\tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) \pm
\tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:

@@

\textit{If we define the residual \(\rho(z)\) as the difference
  between the expected effect at \(x_s\) and the region-effect, i.e
  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the region-variance
  \(\sigma^2(z_1, z_2)\) equals to the integration of the point-wise
  variances plus the squared residuals inside the interval:
\begin{equation}
 \sigma^2(z_1, z_2) = \frac{\int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z}{z_2 - z_1}
\end{equation}}
%

@@

\textit{If we define as \(\rho^2(z_{k-1}, z_{k})\) the
  region-residual, where
  \(\rho^2(z_{k-1}, z_{k}) = \int_{z_{k-1}}^{z_k} \rho^2(z)\partial
  z\), then the two definitions of ARE,
  \(f^{\mathtt{ARE}}, \tilde{f}^{\mathtt{ARE}}\) are equivalent, if
  and only if the sequence of limits \(\mathcal{Z}_s\) are such that
  \(\sum_{k=1}^K \rho^2(z_{k-1}, z_{k}) = 0\).}

@@

Theorem 1 decouples the aggregated uncertainty inside a bin, i.e.,
\(\mathcal{H}_{bin}(z_1, z_2) := \sigma^2(z_1, z_2)\), into two
terms. The first term quantifies the accumulated uncertainty due to
the natural characteristics of the experiment, i.e.,
\(\mathcal{H}(z_1, z_2) = \int_{z_1}^{z_2}\mathcal{H}^2(z) \partial z\),
and the second term adds extra nuisance uncertainty due to limiting
the resolution
\(\mathcal{H}_n(z_1, z_2) := \int_{z_1}^{z_2}\rho^2(z) \partial z
\). Therefore, it holds that:
\begin{equation}
  \label{eq:uncertainty_bin}
  \mathcal{H}_{bin}(z_1, z_2) = \mathcal{H}(z_1, z_2) + \mathcal{H}_n(z_1, z_2)
\end{equation}
%
In Theorem 2 we prove that if we find a sequence of bins, where each
bin adds no nuisance uncertainty, then we will get exactly the same
ARE explanation as in the original definition. In general, we use the
quantity \(Q_{\mathcal{Z}} = \sum_{k=1}^K \rho^2(z_{k-1}, z_{k})\) to
measure the dissimilarity between the actual definition of ARE
\(f^{\mathtt{ARE}}\) and the region-based definition
\(\tilde{f}^{\mathtt{ARE}}\) using the sequence of intervals
\(\mathcal{Z}_s\). In this way, we access a quantified metric for
evaluating how good is a particular splitting.

@@

% Under this
% assumption, and due to the law of large numbers, the approximations
% become more robust, as the bins become wider. On the other hand, bins
% should group local explanations with similar effects and enlarging a
% bin is under the danger of adding nuisance uncertainty.

Both approximations are unbiased only under the assumption that the
points are uniformly distributed inside the intervals. We set-up the
following optimization problem:

\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \tau_k \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
  & \tau_k = 1 - 0.2 \frac{|S_k|}{N} \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
We search for the sequence of intervals \(z_0, \ldots, z_K\) that
minimizes the sum of bin-costs. The cost of each bin is the mean
variance \(\hat{sigma}^2_k\) scaled by the bin length \(\Delta z_k\)
and discounted by the term \(\tau_k\). The term \(\tau_K\) acts as a
tie-braker in cases of two candidate bins with equal mean variance,
favoring the selection of the bigger bin. The constraint of at least
\(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
\textit{robust} estimation.

@@

The simulation examples, where the data-generating distribution
\(p(\mathbf{X})\) and the predictive function \(f(\cdot)\) are defined
by us, enable the evaluation of competitive approaches against a solid
ground-truth. We follow this common XAI
practice~\citep{aas2021explaining, herbinger2022repid} for providing
secure empirical evidence about the superiority of ARE method.
